The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-02 thesis/extra_rv_models.tex
discussion revisions

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Scl and UMi. Our goal is to derive the systemic
velocity and velocity disperions, and understand if the galaxies show
evidence of tidal features, e.g.~a velocity gradient or a rising outer
velocity dispersion. Our derived velocity distributions are consistent
with literature. We find weak evidence for a velocity gradient in Scl
and a rising outer velocity dispersion. However, Scl's velocity gradient
is misaligned with its proper motion, likely inconsistent with a tidal
interpretation. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
observable features of tidal disruption given current velocity
observations.

@@

For both Sculptor and Ursa Minor, we construct literature samples of
radial velocity measurements. We combine these samples with J+24's
Bayesian likelihoods to produce likely members using additional RV
information.

@@

We remove stars with significant velocity dispersions as measured
between observations in a study or between studies, to filter out
unreliable measurements or possible spectroscopic binaries. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically)..

Next, we need to correct the coordinate frames for the solar motion and
on-sky size of the galaxy. We transform the frame into the galactic
standard of rest (GSR). The next step is to account for the slight
differences in the direction of each radial velocity. Let the \(\hat z\)
be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
the angular distance between the centre of the galaxy and the individual
star, the corrected radial velocity is then \begin{equation}{
v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the line of
sight velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
tangental velocities in RA and Dec, and \(\theta\) is the position angle
of the star with respect to the centre of the dwarf. The correction from
both effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We then use the \(v_z\) values for the
following modelling, however repeating with uncorrected, heliocentric
velocities does not significantly affect the results.

The combined likelihood, including RV information, becomes
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
}\end{equation} where we assume that the satellite and background
distributions are Gaussian. Specifically, \begin{equation}{
\begin{split}
{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
dominates the uncertainty budget. We assume a halo/background velocity
dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

@@

We use Monte-Carlo Markov chains (MCMC) to calculate the posterior
distributions of our models of the intrinsic velocity dispersion and
possible velocity gradients in Scl and UMi.

@@

For all models, we ado \begin{equation}{
\mu_{0} \sim N(0\,\kms, \sigma^2_{\rm halo})\\
\sigma_{0} \sim U(0\,\kms, 20\,\kms),
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above, a reasonable
assumption for dwarfs in orbit around the MW.

For the simplest model, \(\mu\) and \(\sigma\) are held
fixed---\(\mu(\xi, \eta) = \mu_0\) and \(\sigma(\xi, \eta) = \sigma_0\).
There are no additional parameters besides \(\mu_0\) and \(\sigma_0\).

To model a velocity gradient, we add two parameters, the gradient
\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta\\
\sigma = \sigma_0
}\end{equation}

To model an increasing velocity dispersion, we add a third parameter,
\(c\) which applies as \begin{equation}{
\log \sigma(R_{\rm ell}) = \log \sigma_0 + c\,\log\left(R_{\rm ell} / R_h\right) \\
\mu = \mu_0 + a\,\xi + b\,\eta\\
}\end{equation} with \(c \sim U(0, 1)\) where \(R_{\rm ell}\) is the
elliptical radius. We only consider this model including the velocity
gradient as a velocity gradient is the simplest way to increase the
dispersion gradient.

Bayes factors quantify the relative evidence between Bayesian model. If
the Bayes factor

Our models are \emph{nested}, where the fiducial model (\(\sigma_0\) and
\(\mu_0\) only) always arises from setting parameters of the more
complex models to zero (\(a=0, b=0\) and/or \(c=0\) ). As a result, we
can use the Savage-Dickey method to calculate the Bayes factors
\citep{dickey+lientz1970}, where the Bayes factor is then the relative
density of posterior versus prior samples where \(a=b=c=0\). To
calculate densities, we use a Silverman-bandwidth KDE to smooth the
samples.

@@

\begin{table*}[t]
\centering
\caption[Ursa Minor RV fits]{MCMC fits for UMi velocity dispersion. }
\label{tbl:umi_rv_mcmc}
\begin{tabular}{lllll}
\toprule
study & mean & sigma & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
\midrule
all & $-245.8\pm0.3$ & $8.8\pm0.2$ & +1.3 & +0.9\\
pace & $-244.6\pm0.4$ & $9.0\pm0.3$ & +0.3 & +0.5\\
spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ & +1.8 & -0.06\\
apogee & $-245.6\pm1.2$ & $10.0_{-0.8}^{+1.0}$ & +1.0 & +0.5\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_vel_gradient_scatter.pdf}}
\caption[Scl velocity gradient]{The corrected LOS velocity along the
best fit rotational axis. RV members are black points, the systematic
\(v_z\) is the horizontal grey line, blue lines represent the
(projected) gradient from MCMC samples, and the orange line is a rolling
median (with a window size of
50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter.pdf}}
\caption[Sculptor's velocity gradient orientation]{RV members of
Sculptor plotted in the tangent plane coloured by corrected velocity
difference from mean \(v_z - \bar v_z\) . The black ellipse marks the
half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
arrows mark the proper motion (PM, GSR frame) and derived velocity
gradient (rot) vectors (to scale).}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/sigma_v_gradient.pdf}}
\caption[Possible gradients in the velocity dispersion]{The observed
velocity dispersion gradient (black) in 10 equal number bins and the
derived slopes from the model fitting. Scl shows moderate evidence for
an increasing velocity dispersion with radius. UMi's evidence for a
radial gradient is weaker.}
\end{figure}

>>>>

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Sculptor (Scl) and Ursa Minor (UMi). Our goal is to
understand if the galaxies show evidence of tidal features---for
example, a velocity gradient or rising outer velocity dispersion. Our
derived systemic velocities and dispersions are consistent with
literature. We find weak evidence for a velocity gradient in Scl and a
rising outer velocity dispersion. However, Scl's velocity gradient is
misaligned with its proper motion, likely inconsistent with a tidal
interpretation. Scl's velocity dispersion also mostly rises within the
effective radius, more likely a feature of inner structure than tidal
disruption. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
clear features of tidal disruption given current velocity observations.

@@

For both Sculptor and Ursa Minor, we construct literature samples of LOS
velocity measurements. We combine these samples with J+24's Bayesian
likelihoods to produce LOS velocity-informed membership probabilities.
We select stars with moderate to large membership probabilities for
future analysis.

@@

We next remove stars with significant velocity dispersions as measured
between observations in a study or between studies, to filter out
unreliable measurements or possible spectroscopic binaries. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (as a star
typically fewer than four velocity measurements).

We correct the velocities for the solar motion and on-sky size of the
galaxy. First, we transform the velocities into the galactic standard of
rest (GSR), a reference frame centred at the Sun's position but at rest
with respect to the Galactic centre. Next, we transform velocities to
correct for the apparent gradient induced by the dwarf's proper motion.
Let the \(\hat z\) be the direction from the sun to the dwarf galaxy.
Then if \(\phi\) is the angular distance between the centre of the
galaxy and the individual star, the corrected radial velocity is then
\begin{equation}{
v_{\rm gsr}' = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
}\end{equation} \{eq:v\_z\} where \(v_{\rm los, gsr}\) is the LOS
velocity in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the
tangental velocities in RA and Dec in the GSR frame, and \(\theta\) is
the position angle of the star with respect to the centre of the dwarf.
The correction from both effects induces an apparent gradient of about
\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We use the \(v_z\) values for the following
modelling.

We build on J+24's likelihood framework to add velocity information (see
Section~\ref{sec:extra_density}). The total likelihood of a star
belonging to the satellite or foreground/background population is
\begin{equation}{
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm los},
}\end{equation} where we have added \({\cal L}_{\rm los}\), LOS-velocity
likelihood term. We assume that the satellite and background
\(v_{\rm los, gsr}\) distributions are Gaussian. Specifically,
\begin{equation}{
\begin{split}
{\cal L}_{\rm los, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm los, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
}\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
dominates the uncertainty budget. We assume a halo/background velocity
dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.

@@

To model for possible gradients. We use Monte-Carlo Markov chains (MCMC)
to calculate the posterior distributions of our models of the intrinsic
velocity dispersion and possible velocity gradients in Scl and UMi.

@@

We consider a single model designed to derive systemic velocity and the
velocity dispersion, detect velocity gradients, and detect a rising
velocity dispersion. The mean velocity at a given point on the sky,
\(\mu\), is assumed to be, \begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta
}\end{equation}

for tangent-plane coordinates \(\xi\) and \(\eta\), systemtic velocity
\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
dispersion at a given position, \(\sigma\), is assumed to depend as a
power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
at \(R_h\), and \(c\) is the velocity dispersion gradient slope.

Our model has parameters with priors \begin{equation}{
\mu_{0} \sim N(0, \sigma^2_{\rm halo}) \,\kms\\
\sigma_{0} \sim U(0, 20)\ \kms \\
a \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
b \sim N(0, 6^2)\ \kms\,{\rm deg}^{-1}\\
c\sim N(0, 0.3^2)\ {\rm dex} \\
}\end{equation} where \(\sigma_{\rm halo} = 100\,{\rm km\,s^{-1}}\) is
the velocity dispersion of the MW halo adopted above,
\(N(\mu, \sigma^2)\) is a standard normal distribution with mean \(\mu\)
and variance \(\sigma^2\), and \(U(l, h)\) is a uniform distribution
between \(l\) and \(h\).

Bayes factors quantify the relative evidence between Bayesian models. We
can use Bayes factors to compare the above model to simpler models
without gradients in \(\sigma\) or \(\mu\). Assuming that a
gradient-free model represents the case \(a=b=0\) (gradient free) or
\(c=0\) (fixed dispersion) models, we can quickly calculate the relative
Bayes factors without running another model. We use the Savage-Dickey
method \citep{dickey+lientz1970}, where the Bayes factor is then the
relative density of posterior versus prior samples when \(a=b=0\)
(gradient free) or \(c=0\) (fixed dispersion). In order to calculate
these densities, we use a kernel density estimator on the posterior
samples with a Silvermann bandwidth in each dimension.

While this is a more complex model, we find each parameter agrees with
models of reduced complexity.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
\caption[Scl velocity gradient]{\textbf{Top} members of Sculptor plotted
in the tangent plane coloured by corrected velocity difference from mean
\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
scale). \textbf{Bottom}: The corrected LOS velocity along the best fit
rotational axis. RV members are black points, the systematic \(v_z\) is
the horizontal grey line, blue lines represent the (projected) gradient
from MCMC samples, and the orange line is a rolling median (with a
window size of 50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/sigma_v_gradient.pdf}}
\caption[Possible gradients in the velocity dispersion]{The observed
velocity dispersion gradient (black) in 10 equal number bins and the
derived slopes from the model fitting. Scl shows moderate evidence for
an increasing velocity dispersion with radius. UMi's evidence for a
radial gradient is weaker, and Scl's velocity dispersion appears to be
flat outside of \(R_h\).}
\end{figure}
